The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-11 rational.tex
conclusion

We have extended the language {{langRev}} that expressed computation
in the commutative semiring of whole numbers to {{langRevEE}} that
expresses computation in the fields of rationals.  Every algebraic
identity that holds for the rational numbers corresponds to a type
isomorphism in our model.  In the process, we have examined the two
function spaces that these give rise to an have seen non-trivial
constructions such as a novel SAT-solver that relies on a
multiplicative trace.

In another sense however, this paper is about the nature of duality in
computation. The concept of duality is deep and significant and in
many cases -- probably most famously in the divide between classical
and intuinionistic mathematics -- contain non-trivial assumptions
about our world-view.

The problem of duality in computation and logic, has many facets of
which `continuation' and `higher order functions' are only one.

\paragraph*{GoI}
The other popular model of computing with values traveling along wire
is GoI -- specifically Mackie's work. How are we related to this
model? Preliminary investogationb suggests that many of the no-cloning
based constructions can be simulated in {{langRevEE}} as well. If
there is a concrete connectoion, then by Curry-Howard the logical
interpretation of {{langRevEE}} should a logic much like LL with a
strong notion of resource preservation, but also with a strong
computational basis.

\paragraph*{Int Construction}
How are we preserving two compact closed structures -- what does this
imply about the Int construction.

\paragraph*{QC} 
A \emph{time traveling} intuition is also applicable here.  In a
normal cicruit, as computational steps are taken values flow from left
to right, i.e. computational time progresses from left to right.  The
action of {{eps+}} causes values, aka information particles, to flow
from right to left i.e. backwards in computational time.  The
interesting thing about this interpretation is that particles that
travel backwards in time get to see and interact with the history of
particles that they coexist with. This gives us an intuitive
interpretation of the isomorphism {{(-b1)*b2<->-(b1*b2)}} (see
Sec. \ref{sec:specific-constructions}) where it would otherwise seem
that {{-b1}} and {{b2}} move in opposite directions. The backward flow
of the {{-b1}} allows it to `see the past' of {{b2}} and is thus
equivalent to both particles moving backward in time.

%% Parallel computing with constraints and backtracking -- breaking
%% dependancies,

@@

%% Partial isomorphisms algebraics
%% do have computational

%% While this equation does indeed have an irrational imaginary root,
%% as we will see -- {{langRevEE}} does not have the means of
%% expressing meaningful computation that uses the root.

The famous example of isomorphism on algebraics is the embedding of
seven binary trees in one. In Figure 1 of Fiore's paper
\cite{Fiore:2004}, he demonstrates this isomorphism by the
corrsponding algebraic manipulation of the types.  The algebraic
manipulation can be viewed as a {{langRevEE}} combinator and it indeed
does constructively witness the isomorphism, i.e., you can run it and
view one data structure encoded by the other. In this context,
consider the following algebraically valid proof of the same
isomorphism:

%% Previously the most famous example of a similar nature is the puzzling
%% isomorphism that one can establish between seven binary trees and one.
%% A binary tree is defined by the datatype
%% \[
%% x = 1 + x * x 
%% \]
%% which can be rearranged to the polynomial equation $x^2 - x + 1 = 0$. By
%% algebraic manipulation we can reason as follows:
\[\begin{array}{rclclclcl}
x^3 &=& x^2 x &=& (x-1) x &=& x^2 - x &=& -1 \\
x^6 &=& 1 \\
x^7 &=& x^6 x &=& x
\end{array}\]
Fiore poses the question of why such algebraic manipulation would make
sense type theoretically, even though some of the intermediate steps
make no sense.  Stated in terms of {{langRevEE}} computations, this
manipulation has no meaningful computational content -- it is always
non-terminating. It suggests that some algebraic manipulations are
somehow more `constructive' than others or dually that {{langRevEE}}
with recursive types somehow lacks the ability to express all
computations that are otherwise algebraically meaningful.

@@

%% -- WAVEFRONT START --

%% But the interesting solution of trees is precisely that they involve
%% irrational numbers. Our usual notation for irrationals such as
%% {{sqrt(2) }} can actually be viewed as a thunk representing the
%% infinite computation {{1.4142..}}.  Irrationals in this sense have a
%% very different information theoretic interpretation from full real
%% numbers.  With Chaitin's $\Omega$ numbers
%% \cite{Chaitin:1975:TPS:321892.321894} he postulates the existence of
%% real numbers that have an infinite amount of information -- ones for
%% example are the solution to the question about the halting properties
%% of every Turning machine.

%% That introduces the natural numbers. No solution for this polynomial
%% over the algebraic numbers. We must extend the numbers with $\omega$
%% to get a solution. We reject this in this paper and prefer to stick to
%% algebraic numbers. The advantage is that all the isomorphisms are
%% valid numerically. With the above type we could subtract $b$ from both
%% sides to show that $0=1$ which is nonsense. We can however have
%% infinite types as long as they have algebraic solutions.

%% In our previous work we introduce recursive types and trace
%% operators. This is dangerous here because infinite loops allow us to
%% prolong paying the debt for as long as we want. The algebraic
%% $\lambda$-calculus~\cite{Vaux:2009:ALC:1630585.1630590} shows that
%% extreme care is needed when combining negative and fractional types
%% with the full $\lambda$-calculus (including recursion).

%% -- WAVEFRONT STOP --

\paragraph*{Summary}
We have contributed an elegant and natural notion of duality that
arises from the monoidal structure of sums and pairs. This notion of
duality has a crisp semantics, clear computational interpretations and
an information theoretic basis.

@@

%% Jacques Carette, Michael Adams, Khaled

>>>>

We have extended the language {{langRev}} that expressed computation
in the commutative semiring of whole numbers to {{langRevEE}} that
expresses computation in the fields of rationals.  Every algebraic
identity that holds for the rational numbers corresponds to a type
isomorphism with a computational interpretation in our model. We have
examined the two function spaces that arise in this model and
developed non-trivial constructions such as a SAT-solver that relies
on a multiplicative trace.

In another sense however, this paper is about the nature of duality in
computation. The concept of duality is deep and significant and in
many cases -- probably most famously in the divide between classical
and intuinionistic mathematics -- contain non-trivial assumptions
about our world-view.  The problem of duality in computation and
logic, has many facets of which `continuation' and `higher order
functions' are only one.  We have contributed an elegant and natural
notion of duality that arises from the monoidal structure of sums and
pairs. This notion of duality has a crisp semantics, clear
computational interpretations and an information theoretic basis.
Before we conclude, we briefly mention some closely related ideas that
warrant further study.

%% \paragraph*{Int Construction} How are we preserving two compact
%% closed structures -- what does this imply about the Int
%% construction.

\paragraph*{Geometry of Interaction.}
GoI was developed by Girard \cite{girard1989geometry}as basis for the
proof theory of Linear Logic.  GoI however has been developed into a
reversible model of computing
\cite{Mackie2011,DBLP:conf/popl/Mackie95,Abramsky:1994:NFG:184662.184664}
in which values traveling along wires, much like {{langRevEE}}.
Preliminary investigations suggest that many GoI machine constructions
can be simulated in {{langRevEE}} by treating Mackie's bi-directional
wires as pairs of wires in {{langRevEE}} and replacing the machine's
global state with a typed value on the wire that captures the
appropriate state.  However a concrete connection between GoI machines
and GoI in general requires further work. This connection is exciting
because when viewed through a Curry-Howard lens it suggests that the
logical interpretation of {{langRevEE}} should a logic much like LL
with a strong notion of resource preservation, but also with a strong
computational basis.

@@

The famous example of isomorphism on algebraics is the embedding of
seven binary trees in one. In Figure 1 of Fiore's paper
\cite{Fiore:2004}, he demonstrates this isomorphism by the
corrsponding algebraic manipulation of the types.  The algebraic
manipulation can be viewed as a {{langRevEE}} combinator and it indeed
does constructively witness the isomorphism, i.e., you can run it and
view one data structure encoded by the other. In this context,
consider the following algebraically valid proof of the same
isomorphism:
\[\begin{array}{rclclclcl}
x^3 &=& x^2 x &=& (x-1) x &=& x^2 - x &=& -1 \\
x^6 &=& 1 \\
x^7 &=& x^6 x &=& x
\end{array}\]
Fiore poses the question of why such algebraic manipulation would make
sense type theoretically, even though some of the intermediate steps
make no sense.  Stated in terms of {{langRevEE}} computations, this
manipulation has no meaningful computational content -- it is always
non-terminating. It suggests that some algebraic manipulations are
somehow more `constructive' than others or dually that {{langRevEE}}
with recursive types somehow lacks the ability to express all
computations that are otherwise algebraically meaningful.

@@

\paragraph*{Quantum Computing} 
A \emph{time traveling} intuition is also applicable here.  In a
normal cicruit, as computational steps are taken values flow from left
to right, i.e. computational time progresses from left to right.  The
action of {{eps+}} causes values, aka information particles, to flow
from right to left i.e. backwards in computational time.  The
interesting thing about this interpretation is that particles that
travel backwards in time get to see and interact with the history of
particles that they coexist with. This gives us an intuitive
interpretation of the isomorphism {{(-b1)*b2<->-(b1*b2)}} (see
Sec. \ref{sec:specific-constructions}) where it would otherwise seem
that {{-b1}} and {{b2}} move in opposite directions. The backward flow
of the {{-b1}} allows it to `see the past' of {{b2}} and is thus
equivalent to both particles moving backward in time.

@@

%% Jacques Carette, Michael Adams, Khaled, Will Byrd, Yin Wang
